Notes: New gaugings in 3d from three-form

### 2023-06-21 Draft_3dH3.tex
Version sent to arXiv. Corrected some typos, removed the H3 with zeta but no Omega.

The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensional external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependence of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0\,,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0\,,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = \tfrac{1}{60}\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}\,.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embeded in $Y^{M}$.

The theory describes the dynamics of the following bosonic fields:
\begin{equation}
	\left\{g_{\mu\nu}, \M_{MN}, A_{\mu}{}^{M}, B_{\mu\,M}\right\}\,,
\end{equation}
with $g_{\mu\nu}$ the 3-dimensinal external metric, $\M_{MN}$ the generalized metric parametrizing the coset space $\EE/\SO{16}$ and the gauge fields $A_{\mu}{}^{M}$ and $B_{\mu\,M}$. It is a gauge theory, invariant under the generalized Lie derivative of parameters $\Upsilon=(\Lambda^{M},\Sigma_{M})$, whose action on a vector $V^{M}$ of weight $\lambda$ is given by
\begin{equation} \label{eq:genlie}
	{\cal L}_{\Upsilon} V^{M} = \Lambda^{N}\partial_{N}V^{M}-60\,(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}\,V^{N}\partial_{K}\Lambda^{L} + \lambda\, V^{M}\partial_{N}\Lambda^{N}+f^{MN}{}_{K}\Sigma_{N}V^{K}\,,
\end{equation}
with $(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}=(1/60)\,f^{M}{}_{NP}f^{PK}{}_{L}$ the projector on the adjoint representation. These transformations are well-defined (in particular, they close into an algebra) only if the parameters $\Sigma_{M}$ and the fields $B_{\mu\, M}$ are covariantly constrained: they have to satisfy algebraic constraints similar to eq.~\eqref{eq:sectionconstraints} and be compatibile with the partial derivatives. We require that
\begin{equation} \label{eq:sectionconstraint2}
	\begin{cases}
		\eta^{MN}C_{M}\otimes C'_{N} = 0\,,\\
		f^{MN}{}_{P}\,C_{M}\otimes C'_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}C_{K}\otimes C'_{L} = 0\,,
	\end{cases}
	\forall\  C_{M}, C'_{M} \in \{\partial_{M}, \Sigma_{M}, B_{\mu\,M}\}\,.
\end{equation}

@@

Thus, for a given twist matrix $U_{M}{}^{\fl{M}}$ satisfying the consistency condition~\eqref{eq:GenLeibniz}, the action~\eqref{eq:ExFTaction} reduces to three-dimensional ${\cal N}=16$ gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity, as is clear from its appearance in \eqref{eq:3dCovDeriv} in the gauge-covariant derivative and the gauge algebra of the vector fields. From eq. \eqref{eq:GenLeibniz}, it has the following expression in terms of the twist matrix and $\rho$:
\begin{equation}
	\begin{split}
		X_{\fl{MN}}{}^{\fl{P}}&=-\rho^{-1}\,\Gamma_{\fl{MN}}{}^{\fl{P}} + \rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{L}}\,\Gamma_{\fl{KM}}{}^{\fl{L}}-\frac{1}{60}\,\rho^{-1}\,f^{\fl{PK}}{}_{\fl{N}}\,f_{\fl{ML}}{}^{\fl{Q}}\,\Gamma_{\fl{KQ}}{}^{\fl{L}}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{M}}\,\Gamma_{\fl{RK}}{}^{\fl{R}}+\left(\delta_{\fl{M}}{}^{\fl{K}}\delta_{\fl{N}}{}^{\fl{P}}-\frac{1}{2}\,f_{\fl{M}}{}^{\fl{LK}}f_{\fl{NL}}{}^{\fl{P}}\right)\,\xi_{\fl{K}}\,.
	\end{split}
\end{equation}
Here, we defined the $\EE$ current $\Gamma_{\fl{MN}}{}^{\fl{P}}=\UI_{\fl{M}}{}^{K}\UI_{\fl{N}}{}^{L}\partial_{K}U_{L}{}^{\fl{P}}$ and the trombone gauging
\begin{equation} \label{eq:Trombone}
	\begin{split}
		\xi_{\fl{M}} 
		&= 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1} - \rho^{-1}\, \Gamma_{\ov{N}\ov{M}}{}^{\ov{N}} \,.
	\end{split}
\end{equation}
The embedding tensor is most nicely expressed once projected on the adjoint representation
\begin{equation} \label{eq:symembeddingtensor}
	\begin{split}
		X_{\fl{MN}} = -\frac{1}{60}\,X_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}} &= - 2\,\rho^{-1}\,\Gamma_{(\fl{MN})} - \rho^{-1}\,\Gamma_{\fl{P}(\fl{M}}{}^{\fl{Q}}\,f_{\fl{N})\fl{Q}}{}^{\fl{P}} + \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \\
		&= - 14\,\rho^{-1}\,(\mathbb{P}_{3875})_{\fl{MN}}{}^{\fl{PQ}}\,\Gamma_{\fl{PQ}} - \frac{1}{4}\,\rho^{-1}\,\eta_{\fl{MN}}\,\Gamma_{\fl{P}}{}^{\fl{P}} - \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \,,
	\end{split}
\end{equation}
with the projection of the current $\Gamma_{\fl{MN}} = - \tfrac{1}{60}\,\Gamma_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}}$. In the following, we will focus on gaugings with $\xi_{\fl{M}}=0$.

The three-dimensional action follows from inserting the Ansätze~\eqref{eq:GSSansatz} in the ExFT action~\eqref{eq:ExFTaction}. The expressions of the kinetic and Chern-Simons terms result immediately from eq.~\eqref{eq:GSSexample}:
\begin{equation}
	\begin{split}
		\mathscr{L}_{\rm kin} &= \frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN} \underset{\rm gSS}{=} \frac{\rho^{2}}{240}\,{\cal D}_{\mu}\M_{\fl{MN}}{\cal D}^{\mu}\M^{\fl{MN}}, \\
		\mathscr{L}_{\rm CS} &\underset{\rm gSS}{=} -\rho^{-1}\,\varepsilon^{\mu\nu\rho}\,X_{\fl{MN}}\,\A_{\mu}{}^{\fl{M}} \left(\partial_{\nu}\A_{\rho}{}^{\fl{N}}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\fl{N}}\right)\,.
	\end{split}
\end{equation}

However, the case of the potential term $\mathscr{L}_{\rm int}$, leading to a potential $V$ for the three-dimensional scalars $\M_{\fl{MN}}$, is more subtle. We expect $V$ to be quadratic in the embedding tensor, which makes the identification of $X_{\fl{MN}}$ more complicated. Moreover, the potential of three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor is unknown, with currently the potential only expressed in terms of the fermion shift matrices of the gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. Thus, in the following, we will follow the truncation procedure carefully and thereby construct the potential of the three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor.

@@

Secondly, knowing that the ${\cal N}=16$ potential must be quadratic in $X_{\fl{MN}}$, we consider the most general quadratic function in $X_{\fl{MN}}$ and develop it in \EE currents using eq.~\eqref{eq:symembeddingtensor}. Note that in higher dimensions, there exist only two quadratic combinations of the embedding tensor:
\begin{equation}
	X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{P}}{}^{\fl{N}} \gM^{\fl{MQ}} \qquad \text{and} \qquad X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{R}}{}^{\fl{S}} \gM^{\fl{MQ}} \gM^{\fl{N}\fl{R}} \gM_{\fl{P}\fl{S}} \,,
\end{equation}
or equivalently
\begin{equation}
	X_{\fl{MN}}X_{\fl{PQ}}\,\M^{\fl{MP}}\eta^{\fl{NQ}} \qquad {\rm and} \qquad X_{\fl{MN}}X_{\fl{PQ}}\,\M^{\fl{MP}}\M^{\fl{NQ}}.
\end{equation}
However, in three dimensions, because the vector fields transform in the adjoint representation of $\EE$, we can write two additional terms
\begin{equation} \label{eq:ExtraX2Terms}
	X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MP}} \eta^{\fl{NQ}} \qquad \text{and} \qquad X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MN}} \eta^{\fl{PQ}} \,,
\end{equation}
where the second term corresponds to the square of the singlet part of the embedding tensor.

@@

The construction of the the $\SL{4} \simeq \SO{3,3}$ twist matrix describing the consistent truncation on $S^3$ is detailed in appendix~\ref{s:SLn}. To demonstrate our methodology, we will now show how the parameter $\lambda$ introduced in \cite{Eloy:2021fhc} can be obtained by the twisting procedure described in appendix~\ref{s:SLn}. Thus, we want to consider the $\SO{3,3} \subset \SO{4,4}$ twist matrix corresponding to $S^3$ \eqref{eq:SLnTwist} and add a 3-form flux via a $\SO{4,4}$ twist. We begin by decomposing $\SO{4,4} \rightarrow \SO{3,3} \times \SO{1,1}$, such that
\begin{equation}
	\begin{split}
		\mathbf{8} &\rightarrow \mathbf{6_0} \oplus \mathbf{1_2} \oplus \mathbf{1_{-2}} \,, \\
		\mathbf{28} &\rightarrow \mathbf{15_0} \oplus \mathbf{6_2} \oplus \mathbf{6_{-2}} \oplus \mathbf{1_0} \,.
	\end{split}
\end{equation}
Correspondingly, we write a $\SO{4,4}$ vector $V^M$ as
\begin{equation}
	V^M = \left( V^A,\, V^z,\, V^{\bar{z}} \right) \,,
\end{equation}
where $A = 1, \ldots, 6$ labels the vector of $\SO{3,3}$ and $z$, $\bar{z}$ label the $\mathbf{1_2}$ and $\mathbf{1_{-2}}$, respectively.

We denote by $U_A{}^{\ov{A}}$ the $\SO{3,3}$ twist matrix corresponding to the $S^3$ truncation constructed in \cite{Lee:2014mla,Baguet:2015iou} and obtained from \eqref{eq:SLnFrame}, \eqref{eq:rho}, \eqref{eq:SLnTwist}. Then, we can add a 3-form potential by twisting with the $\SO{4,4}$ generator
\begin{equation} \label{eq:C2S3example}
	(e^C)^{MN} = \eta^{MN} + C_A \left( t^A \right)^{MN} \,,
\end{equation}
with $C_A$ an element of the $\mathbf{6_{2}}$, and $\left( t^A \right)^{MN}$ the generator corresponding to $\SO{4,4}$ whose only non-zero component is
\begin{equation}
	\left(t^A\right)^{Bz} = \eta^{AB} \,.
\end{equation}
Here $\eta_{MN}$ and $\eta_{AB}$ are the $\SO{4,4}$ and $\SO{3,3}$ invariant metrics, respectively, and are used to raise/lower the corresponding vector indices. Decomposing with respect to the geometric $\SL{3} \times \mathbb{R}^+$ of the $S^3$,  the coordinates on $S^3$ live in the $Y^A = \left( y^i,\, y_i \right)$, $i = 1, 2, 3$, while $C_A = \left( C_i,\, C^i \right)$ naturally contains a 2-form $C^i$. The field strength $H_3 = \partial^A C_A = \partial_i C^i$ is a singlet of $\SO{3,3}$ and thus we see that $\rho^{-1} F_{\ov{M}\ov{N}}{}^{\ov{P}}$ in \eqref{eq:FlatF} is constant and we obtain a consistent truncation with a new 3-form flux. Evaluating the twist matrix
\begin{equation}
	U'_M{}^{\ov{M}} = (e^C)_M{}^N  U_N{}^{\ov{M}} \,,
\end{equation}
explicitly using \eqref{eq:C2S3example} and setting $C^i = \rho^{-1}\,\lambda\, \xi^i$, with $\nabla_i \xi^i = 1$, in the notation of \cite{Eloy:2021fhc}, we obtain precisely the twist matrix used in \cite{Eloy:2021fhc} for the consistent truncation of 6-dimensional ${\cal N}=(1,1)$ supergravity on $S^3$ with $H_3$ flux. 
We will now follow this same procedure in the next section to obtain the consistent truncation of IIB supergravity on $S^3 \times S^3 \times S^1$ with 7-form flux.

@@

\subsection{Adding fluxes to other consistent truncations}
We can also try to apply the same procedure we described here to turn on fluxes in other consistent truncations. One set of natural candidates are the $S^1$ reductions of the consistent truncations to dyonically gauged 4-dimensional maximal supergravities \cite{Guarino:2015jca,Inverso:2016eet}. These 4-dimensional truncations arise from compactifications of 10-dimensional supergravity on $S^6$, $S^5\times S^1$, $S^4\times S^2$, with the 10-d theory corresponding to (massive) IIA/IIB when the spheres have even/odd dimensions, respectively. The corresponding twist matrices for $S^6$, $S^5\times S^1$, $S^4\times S^2$ belong to $\SL{8-p} \times \SL{p}$, with $p = 1,\, 2,\, 3$ for the respective cases, and are constructed as in section \ref{s:SLn}.

@@

However, this $\mathbb{R}^+$ generator cannot correspond to a form potential. If it did, there would also have be a corresponding compact generator in the same representation (with possibly different $\mathbb{R}^+$ chargse), just as in \eqref{eq:133DecompSL4SL4}, associated to the ``dual'' potential. Since we only have one $\SL{8-p} \times \SL{p}$ singlet, this cannot be a $p$-form potential and, therefore, turning on new fluxes with legs on $S^1$ will not lead to consistent truncation for $S^{7-p} \times S^{p-1} \times S^1$, $p = 1,\, 2,\, 3$. This analysis also reveals why the $S^3\times S^3 \times S^1$ case is special: the $\mbf{133}$ contains a 4 form of $\SL{8}$, \textit{i.e.} the $\mathbf{70}$, which naturally contains the relevant singlets under $\SL{4}\times \SL{4}$.

@@

Again in Hopf coordinates:

@@

We can also provide an expression for $H_{(3)}$, and we will do so here in the simplified case where $\Omega=\tilde\Omega=0$. It is convenient to first define some forms:

@@

    \Phi_{i} &= \alpha\,\Big(1+e^{2\psi}\left(\cos^{2}(\ts{\theta})\,\ts{\zeta}_{2}^{2}+\sin^{2}(\ts{\theta})\,\ts{\zeta}_{1}^{2}+\zeta_{i}^{2}\right)\Big)\,\d\phi_{i},\\
    \ts{\Phi}_{i} &= \Big(1+e^{2\psi}\left(\cos^{2}(\theta)\,\zeta_{2}^{2}+\sin^{2}(\theta)\,\zeta_{1}^{2}+\ts{\zeta}_{i}^{2}\right)\Big)\,\d\ts{\phi}_{i},\\

>>>>

The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensional external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependence of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0\,,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0\,,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = \tfrac{1}{60}\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}\,.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embedded in $Y^{M}$.

The theory describes the dynamics of the following bosonic fields:
\begin{equation}
	\left\{g_{\mu\nu}, \M_{MN}, A_{\mu}{}^{M}, B_{\mu\,M}\right\}\,,
\end{equation}
with $g_{\mu\nu}$ the 3-dimensional external metric, $\M_{MN}$ the generalized metric parametrizing the coset space $\EE/\SO{16}$ and the gauge fields $A_{\mu}{}^{M}$ and $B_{\mu\,M}$. It is a gauge theory, invariant under the generalized Lie derivative of parameters $\Upsilon=(\Lambda^{M},\Sigma_{M})$, whose action on a vector $V^{M}$ of weight $\lambda$ is given by
\begin{equation} \label{eq:genlie}
	{\cal L}_{\Upsilon} V^{M} = \Lambda^{N}\partial_{N}V^{M}-60\,(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}\,V^{N}\partial_{K}\Lambda^{L} + \lambda\, V^{M}\partial_{N}\Lambda^{N}+f^{MN}{}_{K}\Sigma_{N}V^{K}\,,
\end{equation}
with $(\mathbb{P}_{248})^{M}{}_{N}{}^{K}{}_{L}=(1/60)\,f^{M}{}_{NP}f^{PK}{}_{L}$ the projector on the adjoint representation. These transformations are well-defined (in particular, they close into an algebra) only if the parameters $\Sigma_{M}$ and the fields $B_{\mu\, M}$ are covariantly constrained: they have to satisfy algebraic constraints similar to eq.~\eqref{eq:sectionconstraints} and be compatibile with the partial derivatives. We require that
\begin{equation} \label{eq:sectionconstraint2}
	\begin{cases}
		\eta^{MN}C_{M}\otimes C'_{N} = 0\,,\\
		f^{MN}{}_{P}\,C_{M}\otimes C'_{N} = 0\,,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}C_{K}\otimes C'_{L} = 0\,,
	\end{cases}
	\forall\  C_{M}, C'_{M} \in \{\partial_{M}, \Sigma_{M}, B_{\mu\,M}\}\,.
\end{equation}

@@

Thus, for a given twist matrix $U_{M}{}^{\fl{M}}$ satisfying the consistency condition~\eqref{eq:GenLeibniz}, the action~\eqref{eq:ExFTaction} reduces to three-dimensional ${\cal N}=16$ gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity, as is clear from its appearance in \eqref{eq:3dCovDeriv} in the gauge-covariant derivative and the gauge algebra of the vector fields. From eq. \eqref{eq:GenLeibniz}, it has the following expression in terms of the twist matrix and $\rho$:
\begin{equation}
	\begin{split}
		X_{\fl{MN}}{}^{\fl{P}}&=-\rho^{-1}\,\Gamma_{\fl{MN}}{}^{\fl{P}} + \rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{L}}\,\Gamma_{\fl{KM}}{}^{\fl{L}}-\frac{1}{60}\,\rho^{-1}\,f^{\fl{PK}}{}_{\fl{N}}\,f_{\fl{ML}}{}^{\fl{Q}}\,\Gamma_{\fl{KQ}}{}^{\fl{L}}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}\,f^{\fl{QK}}{}_{\fl{M}}\,\Gamma_{\fl{RK}}{}^{\fl{R}}+\left(\delta_{\fl{M}}{}^{\fl{K}}\delta_{\fl{N}}{}^{\fl{P}}-\frac{1}{2}\,f_{\fl{M}}{}^{\fl{LK}}f_{\fl{NL}}{}^{\fl{P}}\right)\,\xi_{\fl{K}}\,.
	\end{split}
\end{equation}
Here, we defined the $\EE$ current $\Gamma_{\fl{MN}}{}^{\fl{P}}=\UI_{\fl{M}}{}^{K}\UI_{\fl{N}}{}^{L}\partial_{K}U_{L}{}^{\fl{P}}$ and the trombone gauging
\begin{equation} \label{eq:Trombone}
	\begin{split}
		\xi_{\fl{M}} 
		&= 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1} - \rho^{-1}\, \Gamma_{\ov{N}\ov{M}}{}^{\ov{N}} \,.
	\end{split}
\end{equation}
The embedding tensor is most nicely expressed once projected on the adjoint representation:
\begin{equation} \label{eq:symembeddingtensor}
	\begin{split}
		X_{\fl{MN}} = -\frac{1}{60}\,X_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}} &= - 2\,\rho^{-1}\,\Gamma_{(\fl{MN})} - \rho^{-1}\,\Gamma_{\fl{P}(\fl{M}}{}^{\fl{Q}}\,f_{\fl{N})\fl{Q}}{}^{\fl{P}} + \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \\
		&= - 14\,\rho^{-1}\,(\mathbb{P}_{3875})_{\fl{MN}}{}^{\fl{PQ}}\,\Gamma_{\fl{PQ}} - \frac{1}{4}\,\rho^{-1}\,\eta_{\fl{MN}}\,\Gamma_{\fl{P}}{}^{\fl{P}} - \frac12 f_{\ov{M}\ov{N}}{}^{\ov{P}}\, \xi_{\ov{P}} \,,
	\end{split}
\end{equation}
with the projection of the current $\Gamma_{\fl{MN}} = - \tfrac{1}{60}\,\Gamma_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}}$. In the following, we will focus on gaugings with $\xi_{\fl{M}}=0$.

The three-dimensional action follows from inserting the Ansätze~\eqref{eq:GSSansatz} in the ExFT action~\eqref{eq:ExFTaction}. The expressions of the kinetic and Chern-Simons terms result immediately from eq.~\eqref{eq:GSSexample}:
\begin{equation}
	\begin{split}
		\mathscr{L}_{\rm kin} &= \frac{1}{240}\,D_{\mu}\M_{MN}D^{\mu}\M^{MN} \underset{\rm gSS}{=} \frac{\rho^{2}}{240}\,{\cal D}_{\mu}\M_{\fl{MN}}{\cal D}^{\mu}\M^{\fl{MN}}, \\
		\mathscr{L}_{\rm CS} &\underset{\rm gSS}{=} -\rho^{-1}\,\varepsilon^{\mu\nu\rho}\,X_{\fl{MN}}\,\A_{\mu}{}^{\fl{M}} \left(\partial_{\nu}\A_{\rho}{}^{\fl{N}}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\fl{N}}\right)\,.
	\end{split}
\end{equation}
However, the case of the potential term $\mathscr{L}_{\rm int}$, leading to a potential $V$ for the three-dimensional scalars $\M_{\fl{MN}}$, is more subtle. We expect $V$ to be quadratic in the embedding tensor, which makes the identification of $X_{\fl{MN}}$ more complicated. Moreover, the potential of three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor is unknown, with currently the potential only expressed in terms of the fermion shift matrices of the gauged supergravity \cite{Nicolai:2000sc,Nicolai:2001sv}. Thus, in the following, we will follow the truncation procedure carefully and thereby construct the potential of the three-dimensional ${\cal N}=16$ gauged supergravity in terms of the embedding tensor.

@@

Secondly, knowing that the ${\cal N}=16$ potential must be quadratic in $X_{\fl{MN}}$, we consider the most general quadratic function in $X_{\fl{MN}}$ and develop it in \EE currents using eq.~\eqref{eq:symembeddingtensor}. Note that in higher dimensions, there exist only two quadratic combinations of the embedding tensor:
\begin{equation}
	X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{P}}{}^{\fl{N}} \gM^{\fl{MQ}} \qquad \text{and} \qquad X_{\fl{M}\fl{N}}{}^{\fl{P}} X_{\fl{Q}\fl{R}}{}^{\fl{S}} \gM^{\fl{MQ}} \gM^{\fl{N}\fl{R}} \gM_{\fl{P}\fl{S}} \,,
\end{equation}
or equivalently with eq.~\eqref{eq:symembeddingtensor}
\begin{equation}
	X_{\fl{MN}}X_{\fl{PQ}}\,\M^{\fl{MP}}\eta^{\fl{NQ}} \qquad {\rm and} \qquad X_{\fl{MN}}X_{\fl{PQ}}\,\M^{\fl{MP}}\M^{\fl{NQ}}.
\end{equation}
However, in three dimensions, because the vector fields transform in the adjoint representation of $\EE$, we can write two additional terms
\begin{equation} \label{eq:ExtraX2Terms}
	X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MP}} \eta^{\fl{NQ}} \qquad \text{and} \qquad X_{\fl{MN}} X_{\fl{PQ}}\, \eta^{\fl{MN}} \eta^{\fl{PQ}} \,,
\end{equation}
where the second term corresponds to the square of the singlet part of the embedding tensor.

@@

The construction of the the $\SL{4} \simeq \SO{3,3}$ twist matrix describing the consistent truncation on $S^3$ is detailed in appendix~\ref{s:SLn}. To demonstrate our methodology, we will now show how the parameter $\lambda$ introduced in \cite{Eloy:2021fhc} can be obtained by the twisting procedure described above. Thus, we want to consider the $\SO{3,3} \subset \SO{4,4}$ twist matrix corresponding to $S^3$ \eqref{eq:SLnTwist} and add a 3-form flux via a $\SO{4,4}$ twist. We begin by decomposing $\SO{4,4} \rightarrow \SO{3,3} \times \SO{1,1}$, such that
\begin{equation}
	\begin{split}
		\mathbf{8} &\rightarrow \mathbf{6_0} \oplus \mathbf{1_2} \oplus \mathbf{1_{-2}} \,, \\
		\mathbf{28} &\rightarrow \mathbf{15_0} \oplus \mathbf{6_2} \oplus \mathbf{6_{-2}} \oplus \mathbf{1_0} \,.
	\end{split}
\end{equation}
Correspondingly, we write a $\SO{4,4}$ vector $V^M$ as
\begin{equation}
	V^M = \left( V^A,\, V^z,\, V^{\bar{z}} \right) \,,
\end{equation}
where $A = 1, \ldots, 6$ labels the vector of $\SO{3,3}$ and $z$, $\bar{z}$ label the $\mathbf{1_2}$ and $\mathbf{1_{-2}}$, respectively.

We denote by $U_A{}^{\ov{A}}$ the $\SO{3,3}$ twist matrix corresponding to the $S^3$ truncation constructed in \cite{Lee:2014mla,Baguet:2015iou} and obtained from \eqref{eq:SLnFrame}, \eqref{eq:rho}, \eqref{eq:SLnTwist}. Then, we can add a 3-form potential by twisting with the $\SO{4,4}$ generator
\begin{equation} \label{eq:C2S3example}
	(e^C)^{MN} = \eta^{MN} + C_A \left( t^A \right)^{MN} \,,
\end{equation}
with $C_A$ an element of the $\mathbf{6_{2}}$, and $\left( t^A \right)^{MN}$ the generator corresponding to $\SO{4,4}$ whose only non-zero component is
\begin{equation}
	\left(t^A\right)^{Bz} = \eta^{AB} \,.
\end{equation}
Here $\eta_{MN}$ and $\eta_{AB}$ are the $\SO{4,4}$ and $\SO{3,3}$ invariant metrics, respectively, and are used to raise/lower the corresponding vector indices. Decomposing with respect to the geometric $\SL{3} \times \mathbb{R}^+$ of the $S^3$,  the coordinates on $S^3$ live in the $Y^A = \left( y^i,\, y_i \right)$, $i = 1, 2, 3$, while $C_A = \left( C_i,\, C^i \right)$ naturally contains a 2-form $C^i$. The field strength $H_{(3)} = \partial^A C_A = \partial_i C^i$ is a singlet of $\SO{3,3}$ and thus we see that $\rho^{-1} F_{\ov{M}\ov{N}}{}^{\ov{P}}$ in \eqref{eq:FlatF} is constant and we obtain a consistent truncation with a new 3-form flux. Evaluating the twist matrix
\begin{equation}
	U'_M{}^{\ov{M}} = (e^C)_M{}^N  U_N{}^{\ov{M}} \,,
\end{equation}
explicitly using \eqref{eq:C2S3example} and setting $C^i = \rho^{-1}\,\lambda\, \xi^i$, with $\nabla_i \xi^i = 1$, in the notation of \cite{Eloy:2021fhc}, we obtain precisely the twist matrix used in \cite{Eloy:2021fhc} for the consistent truncation of 6-dimensional ${\cal N}=(1,1)$ supergravity on $S^3$ with $H_3$ flux. 
We will now follow this same procedure in the next section to obtain the consistent truncation of IIB supergravity on $S^3 \times S^3 \times S^1$ with 7-form flux.

@@

\subsection{Adding fluxes to other consistent truncations}
We can also try to apply the same procedure we described here to turn on fluxes in other consistent truncations. One set of natural candidates are the $S^1$ reductions of the consistent truncations to dyonically gauged 4-dimensional maximal supergravities \cite{Guarino:2015jca,Inverso:2016eet}. These 4-dimensional truncations arise from compactifications of 10-dimensional supergravity on $S^6$, $S^5\times S^1$, $S^4\times S^2$, with the 10-d theory corresponding to (massive) IIA/IIB when the spheres have even/odd dimensions, respectively. The corresponding twist matrices for $S^6$, $S^5\times S^1$, $S^4\times S^2$ belong to $\SL{8-p} \times \SL{p}$, with $p = 1,\, 2,\, 3$ for the respective cases, and are constructed as in appendix \ref{s:SLn}.

@@

However, this $\mathbb{R}^+$ generator cannot correspond to a form potential. If it did, there would also have been a corresponding compact generator in the same representation (with possibly different $\mathbb{R}^+$ chargse), just as in \eqref{eq:133DecompSL4SL4}, associated to the ``dual'' potential. Since we only have one $\SL{8-p} \times \SL{p}$ singlet, this cannot be a $p$-form potential and, therefore, turning on new fluxes with legs on $S^1$ will not lead to consistent truncation for $S^{7-p} \times S^{p-1} \times S^1$, $p = 1,\, 2,\, 3$. This analysis also reveals why the $S^3\times S^3 \times S^1$ case is special: the $\mbf{133}$ contains a 4-form of $\SL{8}$, \textit{i.e.} the $\mathbf{70}$, which naturally contains the relevant singlets under $\SL{4}\times \SL{4}$.

@@

Restricting to the parameters $(\zeta_i, \ts{\zeta}_i, \Omega, \ts{\Omega}, \psi)$, $i = 1,2$, and using again Hopf coordinates, the metric is

@@

To provide an expression for $H_{(3)}$ it is convenient to first define some forms:

@@

    \Phi_{1} &= \alpha\,\Big(\big(e^{\ts{\Omega}}\cos^{2}(\ts{\theta})+\sin^{2}(\ts{\theta})\big)\big(e^{\Omega}+e^{2\psi}\,\zeta_{1}^{2}\big)+e^{2\psi+\Omega}\left(\ts{\zeta}_{2}^{2}\cos^{2}(\ts{\theta})+e^{\ts{\Omega}}\,\ts{\zeta}_{1}^{2}\sin^{2}(\ts{\theta})\right)\Big)\,\d\phi_{1},\\
